---
title: "Getting started with orbis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with orbis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(orbis)
```

orbis describes a plot with a layered grammar, then compiles that description
to a scene of drawing primitives. Because the scene is resolution independent,
the same plot object can become an interactive figure in a browser or a 600 dpi
image for a journal, with nothing re-specified.

## The grammar

A plot starts with data and a mapping from variables to visual channels. Layers
are added with `+`.

```{r}
p <- orb(mtcars, x = wt, y = mpg, colour = hp, size = disp) +
  orb_points() +
  orb_labs(title = "Fuel economy", x = "Weight (1000 lbs)",
           y = "Miles per gallon", colour = "Horsepower")
p
```

Recognised channels are `x`, `y`, `colour` (or `color`), `fill`, `size`,
`label`, `group` and `tooltip`. A mapping given inside a layer applies to that
layer only.

Continuous variables mapped to colour use a perceptually uniform ramp;
character or factor variables get a qualitative palette and an interactive
legend.

```{r}
orb(mtcars, x = wt, y = mpg, colour = factor(cyl)) +
  orb_points() +
  orb_theme_dark()
```

## Layers

```{r}
set.seed(1)
df <- data.frame(t = 1:80, v = cumsum(rnorm(80, 0.05)))

orb(df, x = t, y = v) +
  orb_area(alpha = 0.3) +
  orb_line(width = 2, smooth = TRUE) +
  orb_labs(title = "A smoothed series")
```

`orb_bars()` accepts a discrete `x`:

```{r}
counts <- data.frame(g = c("alpha", "beta", "gamma", "delta"),
                     v = c(23, 41, 17, 35))
orb(counts, x = g, y = v, fill = g) +
  orb_bars() +
  orb_options(legend = FALSE, grid = "y") +
  orb_theme_ink()
```

## Small multiples

`orb_facet()` draws one panel per level of a variable. Panels share their scales
by default, which is what makes them comparable at a glance.

```{r}
orb(mtcars, x = wt, y = mpg, colour = hp) +
  orb_points() +
  orb_facet(cyl) +
  orb_labs(title = "By cylinder count", x = "Weight", y = "MPG")
```

Use `scales = "free"` (or `"free_x"`, `"free_y"`) when panels cover very
different ranges, and `ncol` to control the grid:

```{r}
set.seed(3)
d <- data.frame(t = rep(1:40, 3),
                v = c(cumsum(rnorm(40)), cumsum(rnorm(40, 0.3)),
                      cumsum(rnorm(40, -0.2))),
                region = rep(c("north", "south", "east"), each = 40))
orb(d, x = t, y = v) +
  orb_area(alpha = 0.3) + orb_line() +
  orb_facet(region, ncol = 3, scales = "free_y")
```

Colours and sizes are resolved across the whole data set rather than per panel,
so a colour means the same thing everywhere.

## Scales

Scales set palettes, limits and axis transformations.

```{r}
orb(mtcars, x = wt, y = mpg, colour = hp) +
  orb_points() +
  orb_scale_colour("magma", reverse = TRUE) +
  orb_scale_size(range = c(3, 16)) +
  orb_scale_y(limits = c(10, 35))
```

Available palettes:

```{r}
orb_palettes()
```

## Maps

A simplified world dataset ships with the package.

```{r}
str(world_map)
```

`orb_worldmap()` is the quickest way to a map. Supply `values` to shade regions
by a variable.

```{r}
vals <- data.frame(region = c("Brazil", "India", "France", "China", "Nigeria"),
                   v = c(3, 9, 5, 8, 6))
orb_worldmap(values = vals, value_col = "v", projection = "robinson") +
  orb_labs(title = "A small choropleth")
```

Five projections are built in. The orthographic one draws a globe and hides the
hemisphere facing away from the viewer.

```{r}
orb_worldmap(projection = "orthographic", ocean = "#0B1F33") +
  orb_coord_map("orthographic", centre = c(20, 15)) +
  orb_theme_dark()
```

Points are placed on the active projection with `orb_geo_points()`:

```{r}
cities <- data.frame(
  long = c(2.35, -74.0, 151.2, 77.2, -43.2),
  lat  = c(48.86, 40.7, -33.87, 28.6, -22.9),
  pop  = c(11, 19, 5, 32, 13)
)
orb(cities, x = long, y = lat, size = pop) +
  orb_map(fill = "#E9ECEF") +
  orb_geo_points(colour = "#F76707") +
  orb_coord_map("robinson") +
  orb_labs(title = "Cities")
```

## Output

One plot object, several destinations. The format follows the file extension.

```{r, eval = FALSE}
orb_save(p, "figure.png", dpi = 600)   # high-resolution raster for print
orb_save(p, "figure.pdf")              # vector
orb_save(p, "figure.svg")              # vector, still interactive
orb_save(p, "figure.html")             # self-contained interactive page
orb_interactive(p)                     # open in the viewer
```

Raster output is laid out at `width` x `height` logical pixels and then
rendered at the requested `dpi`, so a 300 or 600 dpi figure keeps the same
proportions as what you saw on screen.

The interactive output supports:

* hovering a mark for a tooltip,
* scrolling to zoom and dragging to pan,
* double-clicking to reset the view,
* clicking a legend key to hide or show that series.

Set `orb_options(interactive = FALSE)` if a document must not contain scripts.

## What orbis does not do yet

There is no faceting and there are no statistical layers, so summarise data
before plotting. Text is measured approximately when laying out axes and
legends. The bundled world map is simplified for figure-sized output rather
than for detailed cartography.
